Query Optimization for Dynamic Imputation

### 2017-05-12 rebuttal.tex

    \resp{Agreed. Our database stores tables in files, so our plans are I/O bound, but a database which uses an in-memory representation will need a different Time() function. We have clarified this in the paper.}

  \item S3.4.4: "Drop is a special case of a sequential scan, and its time complexity can be expressed as a function of the number of heap pages read and the I/O cost per page." seems to suggest that drop is implemented as a blocking operator. Why not just take the same approach as selection and treat it as having negligible cost?
    \resp{We mistakenly suggest in S3 that we treat \textit{Drop} as a blocking operator. In
      our implementation, it has $0$ cost, and we have updated the text to reflect that.}

  \item S4: All of the workload queries are (data-size-independent) AVG queries. What about other types of aggregates (e.g., SUM, COUNT)?

    \resp{Though we do not show this explicitly, we repeat all of the queries using COUNT
      aggregates in order to compute the ``Count Fraction''. The timing results are
      identical in this case and the number of tuples returned can be seen
      directly in Table 4.}

    \todo{Do we need to address this?:
    For SUM queries, that may depend more directly on the number of rows returned than the
      quality of imputations for any individual row, the effect of different COUNT results
      dominates the computation AVG*COUNT. How do we handle this?}

@@

  \section{Reviewer 2}
  \begin{itemize}
  \item 1) No new imputation method is presented. The model relies on external imputation techniques, and given it's done on-the-fly, presumably imputed values depend on qualifying query records.

    \resp{Our contribution is the planning algorithm which allows imputation to be interleaved into the query plan.
      We do not present a new imputation method, but our technique is agnostic to the particular imputation method used.
      We reduce the overhead of imputation by performing it only on the tuples which are relevant to a particular query, so the imputed results do depend on the records which are relevant to particular query.}

  \item 2) No comparative evaluation against existing imputation/data cleaning techniques.

    \resp{
      We compare our planning technique for on-the-fly imputation to imputation performed on the full dataset, using the same imputation algorithm in both cases.
      We are not aware of another database technology which offers on-the-fly imputation optimization, so there is no direct comparison for the planning algorithm.
      % Our evaluation compares our technique to an imputation of the full dataset. We use the same imputation algorithm for our on-the-fly imputation as for the full dataset imputation. Cleaning the full dataset is the standard method of handling missing values.
    }

@@

  \item D1) The model relies on external imputation techniques, and given it's done on-the-fly, the imputed values will differ based on the qualifying query records?

    \resp{Yes, as discussed in the response to 1.}

  \item D2) The paper does not propose new imputation techniques but rely on existing statistical imputation algorithms. There is no comparative evaluation, nor samples of imputed values to show the goodness of the imputed values. It would be nice to see a comparative evaluation of different statistical imputation methods, and suggestions for extensions that can fit within the proposed imputation model.

    \resp{
      We have added two new imputation methods to the experiments, to show that the planning algorithm can be used successfully with different imputation methods.
      We have also added a subsection to the algorithm section explaining how we chose the cost functions for each imputation method that we test and giving guidelines for choosing cost functions for a new imputation method.
    }
  \end{itemize}

@@

  \item W3. More experiments needed.

  \resp{
  We've extended the evaluation section to include more impute operators and an evaluation of planning times.
  }

  \item D1. Section 3, which is the meat of the paper, is very overloaded, without getting in depth in various parts of the system. It definitely should be split in more sections, and more details should be given. In particular:
    -- Give some more information about cardinality estimation. Also, Algorithm 3 is too far from the text, and needs some explanation too.
    -- Some more details are also required in the Imputation quality section. Why are you choosing these specific formulas for Loss, especially for the impute operator. Some examples would help.
    -- Algorithms 1 and 2 could probably be presented in an easier to grasp way.

    \resp{
    We've broken up this section and consolidated algorithm notation. The cardinality estimation section provides more detail about the algorithm. We also
    added intuition for how the heuristic Penalty (a renaming of Loss, for clarity) and Time should be designed. We walk through these definitions for three different
    impute operators.
    }

@@

  \item D2. It would be nice to show how exactly your approach can be implemented in an existing optimizer. This comes up in various places in the paper:
    -- When defining the search space, why not first define the transformation operators that inject the imputation operators. Then any optimizer can pick these rules up and add them in its plan enumeration. Of course, the used approach that follows System-R's plan enumeration is useful, but why lose generality when presenting the search space. Moreover, showing the transformation rules can help you show the size of the search space that will better justify the restrictions you apply.
    -- Why use simplified cost formulas for the traditional relational operators and not existing systems' formulas?
    -- It would be great (but not necessary of course) to add your techniques in an existing DBMS, such as Postgres to see the impact on a real system in your experiments.

    \resp{
    \todo{this is a bad response...}
      We felt that the overhead of reimplementing an existing cost model would not be worth the benefit.
      However, we added a new section that discusses in detail the main components for \ProjectName{}, so that developers of existing systems can use this as a roadmap
      for implementation.

@@

  \item D5. Can you have plans with a mix of drop and impute operators?

    \resp{Yes.}

  \item D6. During plan enumeration, can you safely prune plans at some point, similar to what a traditional optimizer would do?

    \resp{We maintain the set of Pareto optimal plans at each point in the optimization. This set is significantly smaller than the set of all plans, so it represents some pruning. We found that the size of these Pareto sets is small enough that we did not need to prune further.

      If necessary, it is possible to sample from the Pareto optimal plans during the plan enumeration. The final selected plan may not be Pareto optimal, but this is a potential way to control the number of plans.
      To perform this approximation, we only add plans to the frontier that are non-dominated and are sufficiently different from existing plans in the frontier. The plan already in the frontier should
      provide similar tradeoffs and reduces the search space explored. \ProjectName{} only employs this optimization when a large number of tables requiring imputation are involved in the query.
      }

  \item D7. When defining the cost of impute operator, how do you know a priori the number of iterations for the algorithm (given it has to converge)?
  For the iterative regression impute operatore, we set a predefined number of iterations as an upper bound on its execution. If the imputation converges,
  we can return immediately, otherwise we return after this upper bound is reached. This is a common approach used in iterative optimization techniques, such
  as gradient descent.

@@

  \item D8. Are there impute implementation that are non-blocking? Or do they always have to perform multiple passes over the data?

    \resp{Yes. For example, in time series data, it is reasonable to fill the most recent non-null value forward. This kind of imputation isn't useful for survey data, so we didn't consider it for our experiments. We have extended the experiments section to consider a simple non-block imputation method relevant to our dataset: mean value imputation. This is a common technique, which fills in  missing values
    with the average observed value. In our case, we implement a non-blocking variant that estimates the mean from the plan's histograms, avoiding the need to perform a full pass over
    the data before imputing.}

### 2017-05-12 paper/experiments.tex
paper: pass over experiments section

For each query, we evaluate \todo{update: three different configurations}: \ProjectName{} optimizing for
quality ($\alpha=0$), \ProjectName{} optimizing for performance ($\alpha=1$), and imputing
at the base tables followed by executing the query traditionally.

\Cref{fig:runtimes} shows a summary of the performance results. The quality-optimized
queries are an \todo{update/validate: order-of-magnitude faster than imputing on the base tables. We get another order-of-magnitude speedup when optimizing for performance}. This performance
differential means it is feasible
to explore multiple imputation operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

\begin{figure}
\includegraphics[width=\columnwidth]{figures/running_times_combined_bar.png}
\caption{Query runtimes with imputation on base tables, \ProjectName{}
    optimizing for quality ($\alpha=0$), and \ProjectName{} optimizing for
    performance ($\alpha=1$). Quality-optimized, on-the-fly imputation provides an order of
    magnitude speedup over imputing on base tables. Efficiency-optimized on-the-fly
    imputation can provide another order-of-magnitude speedup at the expense of quality.}
\todo{update with new figure}   
\label{fig:runtimes}
\end{figure}

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error
(SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results when compared to
imputing on the base tables and executing the query on this cleaned base data.
This measure indicates how much error is introduced by on-the-fly imputation compared to
imputation on the base tables. This comparison is the relevant one in practice --- an
analyst would only be considering the trade-offs in imputation quality and time in the first
place for dirty input data, for which ground truth is unavailable.

@@

\todo{Distinguish between underlying imputation methods}
For $\alpha=0$ and $\alpha=1$, we pair up tuples and compute SMAPE for all queries. That is,
we compute tuple-wise absolute percentage deviations within each iteration of a query, and
we report this value averaged over all iterations.  We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
to the base-imputation approach is low---\todo{update: between \lowsmapealphazero{} and
\highsmapealphaoneexacs{} percent}---indicating that on-the-fly imputation produces similar
results to imputation at the base tables. Notably, the results with largest error correspond
most closely to one existing approach of dropping all \nullv{}s.

@@

\begin{table}
\centering
\input{tables/perf_summary.tex}
\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
    parameterizations. To calculate SMAPE, \ProjectName{} results are compared to query results returned
    from executions with base table imputation. Queries optimized
    for quality ($\alpha=0$) generally achieve lower error than queries optimized for
    efficiency ($\alpha=1$). The count fraction column shows the number of tuples used in calculating each aggregate
     as a fraction of the number of tuples used when running the same query after imputing on the base table.
    A lower count share reflects more potential for errors.}
\todo{update with new figure}
\label{table:smape}
\end{table}

\begin{figure}
\centering
\includegraphics[scale=0.4]{figures/pareto_frontiers_plot.png}
\caption{\ProjectName{}'s final Pareto frontier for \Cref{q1} and \Cref{q6}. 
For clarity, we have removed from the plot plans that provide nearly indistinguishable tradeoffs in either dimension. }
\label{fig:pareto-frontiers}
\end{figure}

\subsubsection{Runtime of Base-table Imputation}
In many real-world cases, applying the imputation step at the base table is prohibitively
expensive.
To illustrate the increasing difficulty of such an approach as datasets scale, we run the following query over the ACS dataset:
\begin{lstlisting}[breaklines]
SELECT AVG(c0) FROM acs_dirty;
\end{lstlisting}
Applying the imputation operation to the base table is extremely expensive, as it imputes over all attributes in all rows, in our case
completing in \acsbaseresultminutes{}. In contrast, \ProjectName{} executes a quality-optimized version
in \acsimputedbzeroresult{} and a runtime-optimized version in \acsimputedboneresult{}. This highlights the potential
benefit of using our system for early data exploration. Note that the performance differential relative to base table imputation
would likely widen further if selection predicates were added to the query.

For every imputation inserted into the query plan, a new statistical model is instantiated and trained before being used for imputation.
Although it is tempting to further optimize query execution in \ProjectName{} by pre-training imputation models while the analyst begins to make queries,
this is not desirable because the specific rows of data, set of complete attributes, and set of attributes to impute are unknown until runtime.

\subsubsection{Planning Times}
\begin{table}
\centering
\input{tables/planning_summary_exact_pareto.tex}
\caption{Planning times are sub-second for up to 4 joins. For 5+  joins planning times rise. Many practical queries
can be formulated with a smaller number of joins, but to support more complex queries \ProjectName{} provides an
optimization that maintains approximate Pareto sets for queries with many joins. This reduces the 29 second
planning time shown for 7 joins by approximately 45\%, on average.
\todo{update with new table and values}
}
\label{table:planning-times}
\end{table}

@@

However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
We explored the limits our planning algorithm by constructing a series of queries with increasing number of joins.
We used the CDC tables and evaluated increasing number of joins across tables.
Planning times for queries involving 1--5 joins, which is a practical value for real world exploratory queries, take less than a second.
\todo{update: For queries with 7 joins, planning time increases substantially to 29 seconds, on average. We introduced an optimization that reduces
the plan space by maintaining only approximations of the Pareto set for queries involving a large number of joins.
With this optimization, the same set of queries have an average planning time of 16 seconds (2 seconds SE).}

>>>>

For each query, we evaluate \todo{update: three different configurations}: \ProjectName{} optimizing for quality ($\alpha=0$) and \ProjectName{} optimizing for performance ($\alpha=1$) and as a baseline, we fully impute the tables used by each query then run the query on the imputed tables, simulating the process that an analyst would follow to apply existing imputation techniques.

\Cref{fig:runtimes} shows a summary of the performance results. The quality-optimized
queries are an \todo{update/validate: order-of-magnitude faster than the baseline. We get another order-of-magnitude speedup when optimizing for performance}. This performance
differential means it is feasible
to explore multiple imputation operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of fully imputing the dataset.

\begin{figure}
\includegraphics[width=\columnwidth]{figures/running_times_combined_bar.png}
\caption{Runtimes for the queries in~\Cref{tbl:queries}. $\alpha=0$ (resp. $\alpha=1$) is \ProjectName{} run at the highest (resp. lowest) quality setting. Baseline is full table imputation followed by running the query.}
\todo{update with new figure}   
\label{fig:runtimes}
\end{figure}

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results compared to the baseline.
This measures the error introduced by on-the-fly imputation as compared to full imputation on the base tables.
This comparison is the relevant one in practice --- an analyst would only be considering the trade-offs in imputation quality and time in the first place for dirty input data, for which ground truth is unavailable.

@@

\todo{Distinguish between underlying imputation methods}
For $\alpha=0$ and $\alpha=1$, we pair up tuples and compute SMAPE for all queries. That is,
we compute tuple-wise absolute percentage deviations within each iteration of a query, and
we report this value averaged over all iterations.  We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
to the baseline is low---\todo{update: between \lowsmapealphazero{} and
\highsmapealphaoneexacs{} percent}---indicating that on-the-fly imputation produces results similar to the baseline. Notably, the results with largest error correspond
most closely to one existing approach of dropping all \nullv{}s.

@@

Queries optimized
    for quality ($\alpha=0$) generally achieve lower error than queries optimized for
    efficiency ($\alpha=1$). The count fraction column shows the number of tuples used in calculating each aggregate
     as a fraction of the number of tuples used when running the same query after imputing on the base table.
    A lower count share reflects more potential for errors.

\begin{table}
\centering
\input{tables/perf_summary.tex}
\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$ parameterizations, as compared to the baseline.}
\todo{update with new figure}
\label{table:smape}
\end{table}

\begin{figure}
\centering
\includegraphics[scale=0.4]{figures/pareto_frontiers_plot.png}
\caption{\ProjectName{}'s final Pareto frontier for \Cref{q1} and \Cref{q6}. 
For clarity, we have removed from the plot plans that provide nearly indistinguishable tradeoffs in either dimension.}
\label{fig:pareto-frontiers}
\end{figure}

\subsubsection{Baseline Runtime}
In many real-world cases, applying imputation to an entire dataset is prohibitively expensive.
For example, if we run CART imputation on the full ACS dataset, it completes in \acsbaseresultminutes{}.

In contrast, \ProjectName{} executes the following query:
\begin{lstlisting}[breaklines]
SELECT AVG(c0) FROM acs;
\end{lstlisting}
over the ACS dataset (also using CART imputation) in \acsimputedbzeroresult{} when optimizing for quality and \acsimputedboneresult{} when optimizing for runtime.
The performance difference comes from the fact that \ProjectName{} only needs to perform imputation on the single column required by the query.
Adding selection predicates would further reduce the query runtime by reducing the amount of required imputation.
An analyst could do the same, but tracking the required imputations would get increasingly complicated as the queries became more complex.
This highlights the benefit of using our system for early data exploration.

\todo{Should this be here?}
For every imputation operator inserted into the query plan, a new statistical model is instantiated and trained before being used for imputation.
Although it is tempting to further optimize query execution in \ProjectName{} by pre-training imputation models while the analyst begins to make queries,
this is not desirable because the specific rows of data, set of complete attributes, and set of attributes to impute are unknown until runtime.

\subsubsection{Planning Times}
\begin{table}
\centering
\input{tables/planning_summary_exact_pareto.tex}
\caption{Planning times for varying numbers of joins. \todo{update with new table and values}}
  % Planning times are sub-second for up to 4 joins. For 5+  joins planning times rise. Many practical queries
% can be formulated with a smaller number of joins, but to support more complex queries \ProjectName{} provides an
% optimization that maintains approximate Pareto sets for queries with many joins. This reduces the 29 second
% planning time shown for 7 joins by approximately 45\%, on average.
\label{table:planning-times}
\end{table}

@@

However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
We explored the limits our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
Planning times for queries involving 1--5 joins---a practical value for real world exploratory queries---are less than 1 second.
\todo{update: For queries with 7 joins, planning time increases substantially to 29 seconds, on average.
  It is possible to reduce planning time by approximating the Pareto set, particularly for queries with a large number of joins.
  With this optimization, the queries with 7 joins have an average planning time of 16 seconds (2 seconds SE).}
